---
title: "Iteration with purrr"
format: html
editor_options: 
  chunk_output_type: console
---

The purpose of this lesson is to learn how to iteratively apply functions to all elements contained within an object, such as all columns in a data frame, or all entries in a vector.

The "purrr" R package that we will be using in this lesson is *included* in the tidyverse package.

```{r}
# load the tidyverse and demographics dataset

```




The function that we want to iterate first is the `n_distinct()` function from the tidyverse. 

Recall that to pull up the help page for a function, you can use the following syntax:

```{r}
# pull up the help page for n_distinct

```


`n_distinct()` counts the number of unique values in a vector. 

```{r}
# apply n_distinct to the "gender" column of demographics

```


The code below uses the `map()` function to apply `n_distinct()` to every column of the `demographics` data frame simultaneously:

```{r}
# use map() to apply n_distinct to every column of demographics

```


The output of the `map()` function is a list (more on that in a moment).

The following code shows a simpler example, this time applying the `exp()` function to all entries/elements in a simple numeric vector:

```{r}
# use map() to apply exp() to the vector c(4, 5, 6)

```





### Lists

Lists, like vectors, correspond to a collection of values contained in a single object. 

You can use the `list()` function to define a list, for example:

```{r}
# define a list called my_list with three elements: 1, 4, and 7

```

You can extract elements from a list just as you would from a vector, using the square bracket notation. The code below extracts the third element from `my_list`:

```{r}
# extract the third entry from my_list using []

# ask the class of the object above

```

However, the output above is a list itself.


If you want to extract the actual object/value contained within the third element of the list, then you need to use double square parentheses:

```{r}
# extract the third entry from my_list using [[]]

```



Unlike vectors, lists are not vectorized. 

```{r}
#| error: true

# try to add 1 to my_list

```

Why would we ever prefer lists to vectors? The reason is that lists are infinitely more flexible than vectors. While all entries in a vector must be a *single value* of the *same type*, entries of a list can be anything. 


```{r}
# create a list my_complex_list containing 
# (1) the head of demographics, 
# (2) the value 2, and 
# (3) a vector containing "a" and "b"

```

You can also create a *named* list when defining it as if you are defining argument names:

```{r}
# create a named version of my_complex_list

```



You can then extract entries from the named list using `$` or `[[]]`

```{r}
# extract one of the elements from my_complex_list using $

# extract one of the elements from my_complex_list using [[]]

```







### Exercise

Use `map()` to apply `class()` to every column in the demographics dataset, and extract the class of the `household_income` column.









## Using custom functions in purrr


To iterate using your own custom functions in a purrr `map()` function, you can define your function and provide it in the second argument of `map()`. 

```{r}
# define a function called exp_plus_one() that returns exp(x) + 1

# apply it to every entry in the vector c(1, 4, 5)

```

For simple functions like this, we can define the function inside the `map()` function itself:

```{r}
# apply the function exp(x) + 1 to every entry in the vector c(1, 4, 5)
# using an "anonymous" function

```

However, we can go one step further and forego the `function(x)` part entirely using what I call the "tilde-dot" shorthand syntax. 

Here, we use the tilde `~` symbol to "start" an anonymous function and we use a `.` to represent the argument of our anonymous function. 

```{r}
# apply the function exp(x) + 1 to every entry in the vector c(1, 4, 5)
# using the "tilde-dot" syntax

```


To decide what should go inside the `~{}`, I typically use a representative value to test out my code first. 

Let's count the number of missing values in each column of demographics.

For example, below, I take a single column `demographics$pregnant`, and I write out the code that I want to apply to it:

```{r}
# compute the number of missing values in the "pregnant" column of demographics

```

Generalize this code for the map function:

```{r}
# use a map function to compute the number of missing values in every column

```







### Exercise 

Use the tilde-dot short-hand syntax to compute the number of values in each column that is equal to 1. Recall that if a vector has missing values, and you want to use `sum()`, you will want to provide an argument `na.rm = TRUE` to ignore missing values.










## Alternative map functions for outputting doubles, characters and data frames

Recall our `map()` function previously that we used to count the number of missing values in each column of the demographics data.

```{r}
#| eval: false
map(demographics, ~sum(is.na(.)))
```

A list may not be the most useful format for this information...





### Outputting numeric vectors

The `map_dbl()` function will output a "double" (numeric) vector. 

```{r}
# use map_dbl to count the number of missing values in each column and
# output a numeric vector

```



### Outputting character vectors

The `map_chr()` function can output a *character* vector.

```{r}
# use map_chr to apply class to every column and output a character vector

```




### Outputting data frames

One of the most versatile functions is the `map_df()` function, which outputs a data frame. 


```{r}
# apply map_df to demographics to determine the class of each column, outputting a "wide" data frame

```

If you want your output to be in a "long" format, the function you are applying must output a single-column tibble/data frame.

```{r}
# use tibble() to create a single-column tibble containing the class of demographics$pregnant

# modify this code to be the function call in map_df to create a 
# long-form data frame

# provide an .id argument to include the original column names as a variable

```







### Example

This long-format is very useful if you want to create a plot, such as a bar chart for the number of missing values.

```{r}
# create a bar chart of the number of missing values in each column
# use factors to order the columns by the number of missing values

```

An alternative approach is to `map_df()` is to use `map_dbl()` and `enframe()`

```{r}
# use map_dbl() and enframe() to create a long data frame of the 
# number of missing values in each column

```


